---
title: "continue....."
output:
  html_document: default
  pdf_document: default
bibliography: Andrea.bib
csl: apa.csl
---

```{r setup, results=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = "png")
```


## 7. Data analysis 

First we will load all the libraries that we will need to perform our data analysis:
```{r, results='hide', warning=FALSE, error=FALSE, message=FALSE}
library(tidyverse)
library(caret)
library(ade4) # PCA
library(made4) # PCA
library(scatterplot3d) # PCA
library(admisc) # Read radar objects
library(omicade4) # CCA
library(gbm) # Boosting
library(randomForest) # Random Forest
library(xgboost) # XgBoost
library(data.table) # for setnames (in random forest)
library(DMwR) # SMOTE -> remotes::install_github("cran/DMwR")
```

<br>


#### **Chronic obstructive pulmonary disease (COPD)**

Chronic obstructive pulmonary disease (COPD) is a long-term lung illness in which the lungs' small airways are damaged, making it difficult for air to enter and exit.\

COPD patients commonly experience one or more of the following symptoms:

- having trouble breathing 

- prolonged cough (more than 3 months) 

- a cough with mucus most days

- wheezing (a whistling sound when you breathe)

- chest tightness


Shortness of breath is the most common symptom of COPD. Simple daily tasks like walking short distances or walking a stairway might cause COPD sufferers to become out of breath. Even at rest, breathing might become difficult as the condition progresses.\

Long-term cigarette smoking is the most common cause of COPD. The more cigarettes you smoke, the more likely you are to get COPD.\


<br>


### 7.1 Clinical data cleaning

Before we begin with our analysis, we need to explore our clinical data and clean it.

View data:
```{r}
clinic_data <- read.delim("eclipse_all.txt")
print(as_tibble(clinic_data))
```

Delete irrelevant columns:
```{r}
clinic_data$id <- paste0("0", clinic_data$id) # Add 0 in front of index to match rdr names

clinic_data <- clinic_data[-1] # delete first columns (index 0,1,2...)
row.names(clinic_data) <- clinic_data$id # id as rownames
clinic_data$id <- NULL # delete id colum

clinic_data$SUBJID.t1 <- NULL # index
clinic_data$SCRNDT <- NULL # dates
clinic_data$SUBJID.t3 <- NULL # index
```

Select variables of t1 (data from time 1).
```{r, warning=FALSE}
clinical_data_T1 <- dplyr::select(clinic_data, !ends_with("t3"))
```

Check for missing data:
```{r}
sum(is.na(clinical_data_T1))
```

Delete variables that have more than a 30% of missing values:
```{r}
more_30_missing <- names(clinical_data_T1[ , colMeans(is.na(clinical_data_T1)) > 0.3])
clinical_data_T1 <- clinical_data_T1[, !colnames(clinical_data_T1) %in% more_30_missing]
```

Impute missing values of double variables for the mean value:
```{r}
clinical_data_T1 <- clinical_data_T1 %>%  mutate_if(is.double,
                                                    function(x) ifelse(is.na(x), mean(x, na.rm = T), x)) 
```

Impute missing values of integer variables for the most common value:
```{r}
clinical_data_T1 <- clinical_data_T1 %>%  mutate_if(is.integer,
                                                    function(x) ifelse(is.na(x), which.max(x), x)) 
```

Impute missing values of categroical variables for the most common value:
```{r}
clinical_data_T1[, sapply(clinical_data_T1, function(x) !is.numeric(x))] <- apply(clinical_data_T1[, sapply(clinical_data_T1, function(x) !is.numeric(x))], 
                                                      2, function(x) {x[is.na(x)] <- names(sort(table(x), decreasing = TRUE)[1]); x})
```

Check for missing data:
```{r}
sum(is.na(clinical_data_T1))
```

Characters to factors:
```{r}
ch <- sapply(clinical_data_T1, is.character)
clinical_data_T1[ch] <- lapply(clinical_data_T1[ch], as.factor)
```

Now that we have a clean data set, we will export it:
```{r, eval = FALSE}
write.table(clinical_data_T1, file = "eclipse_all_CLEAN.txt", sep = "\t")
```



<br>



### 7.2 Principal Component Analysis (PCA)

We will first start doing principal component analysis with the radiomic features.\

Recall that PCA is a statistical approach that allows you to summarize the content of big data tables using a smaller set of "summary indices" that are easier to display and study.\


**All clinical variables**

Load the clean clinic data set and convert all columns to be numeric (needed for PCA):
```{r}
clinic_data_clean <- read.delim("eclipse_all_CLEAN.txt")
data <- mutate_all(clinic_data_clean, function(x) as.numeric(as.factor(x)))
data[1:5, 1:8]
```

PCA:
```{r}
results_PCA <- ord(data, type="coa")
plot(results_PCA, classvec = colnames(data), genecol="grey3")
```


<br>



**Differenciate between control (smoker vs non-smoker) and cases (COPD)**

Load data:
```{r}
clinic_data_clean <- read.delim("eclipse_all_CLEAN.txt")
clinic_data_clean <- mutate_all(clinic_data_clean, function(x) as.numeric(as.factor(x)))
data <- as.data.frame(t(clinic_data_clean))
colnames(data) <- rownames(clinic_data_clean) # put id names

## Label for "COPD" or "Non COPD"
class_vector <- as.numeric(data["TRTGRP.t1",])
for(i in 1:length(class_vector)){
  if(class_vector[i] == 1) class_vector[i] <- "COPD"
  else class_vector[i] <- "Non COPD"
}

## Delete TRTGRP.t1 from table
data <- data[!(row.names(data) %in% "TRTGRP.t1"), ]
data[1:5, 1:8]
```

PCA:
```{r}
results_PCA <- ord(data, type="coa")
plot(results_PCA, classvec = class_vector, genecol="grey3")
```

<br>


**Radiomic features: control (smoker vs non-smoker) and cases (COPD)**

Load radiomic features and clinic data. Since they differ in the number of id, we will only keep common ids.
```{r}
## Radar object
file <- listRDA("rdr_L1_Andrea.rda")
rdr_L1 <- file$rdr_L1
df_rdr <- as.data.frame(assay(rdr_L1))

## Clinic data
clean_data <- read.delim("eclipse_all_CLEAN.txt")
clean_data <- mutate_all(clean_data, function(x) as.numeric(as.factor(x))) # convert to numeric
df_clinical <- as.data.frame(t(clean_data)) # id as columns
colnames(df_clinical) <- rownames(clean_data) 

## Keep common ids
id_in_rdr <- colnames(df_rdr) # id of rdr
id_in_clinical <- colnames(df_clinical) # id of clinical
common_id <- Reduce(intersect, list(id_in_rdr,id_in_clinical)) # common id

df_clinical <- select(df_clinical, one_of(common_id))
df_rdr <- select(df_rdr, one_of(common_id))

## Label for "COPD" or "Non COPD"
class_vector <- as.numeric(df_clinical["TRTGRP.t1",])
for(i in 1:length(class_vector)){
  if(class_vector[i] == 1) class_vector[i] <- "COPD"
  else class_vector[i] <- "Non-COPD"
}

df_rdr[1:5, 1:8]
```

PCA:
```{r}
results_PCA <- ord(df_rdr, type="coa")
plot(results_PCA, classvec = class_vector, genecol="grey3")
```




<br>



### 7.2 Canonical correlation analysis (CCA)

Now we will analyze the radiomic features against the variables at time T1 using **canonical correlation**.\

Canonical correlation analysis (CCA) is a statistical technique for extracting information from cross-covariance matrices. If we have two vectors of random variables, X = (X1,..., Xn) and Y = (Y1,..., Ym), and the variables are correlated, canonical-correlation analysis will find linear combinations of X and Y that have the highest correlation. \

Multiple correlation analysis predicts only one dependent variable from several independents, but canonical correlation predicts multiple dependent variables from multiple independents.\

In our case, we will have 2 tables. One with the radiomics features, and another one with the clinical data. We want to have id -> columns and features -> rows.\


Load radiomic features and clinic data. Since they differ in the number of id, we will only keep common ids.
```{r}
## Radar object
file <- listRDA("rdr_L1_Andrea.rda")
rdr_L1 <- file$rdr_L1
df_rdr <- as.data.frame(assay(rdr_L1))

## Clinic data
clean_data <- read.delim("eclipse_all_CLEAN.txt")
clean_data <- mutate_all(clean_data, function(x) as.numeric(as.factor(x))) # convert to numeric
df_clinical <- as.data.frame(t(clean_data)) # id as columns
colnames(df_clinical) <- rownames(clean_data) 

## Keep common ids
id_in_rdr <- colnames(df_rdr) # id of rdr
id_in_clinical <- colnames(df_clinical) # id of clinical
common_id <- Reduce(intersect, list(id_in_rdr,id_in_clinical)) # common id

df_clinical <- select(df_clinical, one_of(common_id))
df_rdr <- select(df_rdr, one_of(common_id))

## Label for "COPD" or "Non COPD"
class_vector <- as.numeric(df_clinical["TRTGRP.t1",])
for(i in 1:length(class_vector)){
  if(class_vector[i] == 1) class_vector[i] <- "COPD"
  else class_vector[i] <- "Non-COPD"
}

df_clinical <- df_clinical[!(row.names(df_clinical) %in% "TRTGRP.t1"), ] # delete TRTGRP.t1 from table

df_clinical[1:5, 1:8]
df_rdr[1:5, 1:8]
```

Combine both data frames into a list.
```{r}
both_tables <- list("rdr" = df_rdr, "clinical" = df_clinical)
sapply(both_tables, dim) # check the dimension are correct
```

Perform CCA:
```{r}
mcoin <- mcia(both_tables, cia.nf=10)
plot(mcoin, axes=1:2, phenovec = class_vector, sample.lab=FALSE, df.color=1:2) # visualize the result
```



<br>



### 7.4 TRTGRP prediction

Now we will use different methods to predict *TRTGRP* with the radiomic features.\

Load clinic data and radiomic features. Since they differ in the number of id, we will only keep common ids.
```{r}
## Radar object
file <- listRDA("rdr_L1_Andrea.rda")
rdr_L1 <- file$rdr_L1
df_rdr <- as.data.frame(assay(rdr_L1))

## Clinic data
clean_data <- read.delim("eclipse_all_CLEAN.txt")
ch <- sapply(clean_data, is.character) # characters to factors
clean_data[ch] <- lapply(clean_data[ch], as.factor)

## Keep common ids
id_in_rdr <- colnames(df_rdr) # id of rdr
id_in_clinical <- rownames(clean_data) # id of clinical
common_id <- Reduce(intersect, list(id_in_rdr,id_in_clinical)) # common id
df_clinical <- clean_data %>% filter(row.names(clean_data) %in% common_id)
df_rdr <- select(df_rdr, one_of(common_id))
```

View a summary of our variable of interest *TRTGRP.t1*:
```{r}
table(clean_data$TRTGRP.t1)
TRTGRP_values <- as.numeric(clean_data$TRTGRP.t1)
table(TRTGRP_values)
```

We have three different levels: COPD Subjects Non-smoker, Controls and Smoker Controls. We will combine it into two levels: COPD (cases) and Non COPD (controls).
```{r}
for(i in 1:length(TRTGRP_values)){
  if(TRTGRP_values[i] == 1) TRTGRP_values[i] <- "COPD"
  else TRTGRP_values[i] <- "Non COPD"
}
table(TRTGRP_values)
TRTGRP_values <- as.numeric(as.factor(TRTGRP_values))
table(TRTGRP_values)
```

We are interested in predicting *TRTGRP.t1* given the radiomic features. So we will add this variable to the table of radiomic features.\


Add the values to the data frame and put id as rows instead of columns:
```{r}
df_rdr2 <- rbind(df_rdr, "TRTGRP_values" = TRTGRP_values)
df_rdr3 <- as.data.frame(t(df_rdr2))
rownames(df_rdr3) <- colnames(df_rdr2) # put id names
print(as_tibble(df_rdr3))
```


With the data we have, we need to balance the data.

Check how much data we have of each type:
```{r, out.width="50%"}
barplot(prop.table(table(TRTGRP_values)), col = rainbow(2), ylim = c(0, 1), main = "TRTGRP_values Distribution")
```

We can clearly see that we don't have balanced data, this will produce not very accurate results in our models.
```{r}
table(TRTGRP_values)
```


<br>



**Boosting**

We create a function that takes as a parameter your entire dataset and returns the accuracy of a boosting model (GBM).\

```{r}
Boosting <- function(data){
  
  #### Select train 70% and test 30% randomly
  rows <- sample(nrow(data), .7 * nrow(data))
  train_data <- data[rows, ] # train data
  test_data <- data[-rows, ] # test data
  
  #### Train a model using our training data
  model_gbm <- gbm(TRTGRP_values ~.,
                data = train_data,
                distribution = "multinomial",
                cv.folds = 10,
                shrinkage = .01,
                n.minobsinnode = 10,
                n.trees = 500)       # 500 tress to be built
  
  #### Make predictions on the test data
  pred_test <- predict.gbm(object = model_gbm,
                   newdata = test_data,
                   n.trees = 500,           # 500 tress to be built
                   type = "response")
  
  #### Create a confusion matrix:
  class_names <- colnames(pred_test)[apply(pred_test, 1, which.max)]
  result <- data.frame(test_data$TRTGRP_values, class_names)
  conf_mat <- confusionMatrix(as.factor(test_data$TRTGRP_values), as.factor(class_names))
  
  return(conf_mat$overall["Accuracy"]) # return Accuracy
}
```

First we will do Boosting *without balancing* the data:
```{r, warning=FALSE}
(boosting_not_balanced <- Boosting(df_rdr3))
```

We get an Accuracy of around 79%.

Now we balance our data with *SMOTE* method, which is an oversampling technique where the  samples are generated for the minority class. 

```{r}
df_rdr3$TRTGRP_values <- as.factor(df_rdr3$TRTGRP_values)
smote_data <- SMOTE(TRTGRP_values ~ ., data  = df_rdr3)                         
table(smote_data$TRTGRP_values) 
```

```{r}
(boosting_smote <- Boosting(smote_data))
```

We get worse results with an Accuracy of 70%.


<br>



**Random Forest**

With Random Forest we can't have columns starting with numbers, so we will change them and add an A before the number:

```{r}
df_rdr4 <- df_rdr3
# Columns that start with numbers
names_number <- colnames(df_rdr4 %>% dplyr:: select(starts_with(c("0","1","2","3","4","5","6","7","8","9"))))
# Add and A before the number
names_changed <- unlist(lapply(names_number, function(x) paste("A", x, sep="")))
# Change the names
setnames(df_rdr4, old=names_number, new = names_changed, skip_absent=TRUE)
```


We create a function that takes as a parameter your entire dataset and returns the accuracy of a Random Forest model.\

```{r}
RandomForest <- function(data){
  
  #### Select train 70% and test 30% randomly
  rows <- sample(nrow(data), .7 * nrow(data))
  train_data <- data[rows, ]
  test_data <- data[-rows, ]
  
  
  #### Train a model using our training data
  model <- randomForest(TRTGRP_values ~ . , data = train_data)
  
  
  #### Make predictions on the test data
  predictions <- predict(model, test_data)
  confusion_matrix <- with(test_data, table(predictions, TRTGRP_values))
  
  
  #### Accuracy of our model
  accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
  return(accuracy)
}
```

First we will do Random Forest *without balancing* the data:
```{r}
(rf_not_balanced <- RandomForest(df_rdr4))
```

We get an Accuracy of around 79%.

```{r}
table(df_rdr4$TRTGRP_values)
```


Now we balance our data with *SMOTE* method, which is an oversampling technique where the  samples are generated for the minority class. 

```{r}
smote_data <- SMOTE(TRTGRP_values ~ ., data  = df_rdr4)                         
table(smote_data$TRTGRP_values) 
```

```{r}
(rf_smote <- RandomForest(smote_data))
```

We get better results with an Accuracy of around 88%.\



<br>




**XGBoost**

We create a function that takes as a parameter your entire dataset and returns the accuracy of a XgBoost model.\

```{r}
XgBoost <- function(data){
  
  #### Select train 70% and test 30% randomly
  rows <- sample(nrow(data), .7 * nrow(data))
  train_data <- data[rows, ] # train data
  test_data <- data[-rows, ] # test data
  
  
  ####  Independent and dependent variables for train and test
  X_train = data.matrix(train_data[,-ncol(data)]) 
  y_train = train_data[,ncol(data)] 
    
  X_test = data.matrix(test_data[,-ncol(data)])  
  y_test = test_data[,ncol(data)] 
  
  
  #### Convert the train and test data into xgboost matrix type
  xgboost_train = xgb.DMatrix(data=X_train, label=y_train)
  xgboost_test = xgb.DMatrix(data=X_test, label=y_test)
  
  
  #### Train a model using our training data
  model <- xgboost(data = xgboost_train,                      
                   max.depth = 3, # max depth 
                   nrounds = 50, # max number of boosting iterations
                   verbose = 0) # omit printing 'train-rmse' for each iteration
  
  
  #### Make predictions on the test data
  pred_test <- predict(model, xgboost_test)
  
  
  #### Convert prediction to factor type:
  pred_test[(pred_test>3)] = 3
  pred_y <- as.factor((levels(y_test))[round(pred_test)])
  
  
  #### Create a confusion matrix:
  conf_mat <- confusionMatrix(y_test, pred_y)
  return(conf_mat$overall["Accuracy"])
  
}
```

First we will do XgBoost *without balancing* the data:
```{r}
(xgboost_not_balanced <- XgBoost(df_rdr3))
```

We get an Accuracy of around 76%.

```{r}
table(df_rdr3$TRTGRP_values)
```


Now we balance our data with *Downsampling* method, which takes the lowest value (in our case Non-smoker Controls = 193) and makes all classes have the same amount of values:
```{r}
down_data <- downSample(x = df_rdr3[, -ncol(df_rdr3)],
                         y = df_rdr3$TRTGRP_values)
table(down_data$Class) 
```

```{r}
(xgboost_down <- XgBoost(down_data))
```

We get worse results with an Accuracy of 60%.

Now we balance our data with *SMOTE* method, which is an oversampling technique where the  samples are generated for the minority class. 

```{r}
smote_data <- SMOTE(TRTGRP_values ~ ., data  = df_rdr3)                         
table(smote_data$TRTGRP_values) 
```

```{r}
(xgboost_smote <- XgBoost(smote_data))
```

We get better results with an Accuracy of 80%.\


Summary:
```{r}
final_results <- data.frame("ACCURACY" = c("No balancing", "SMOTE", "Down Sample"),
                            "Boosting" = c(boosting_not_balanced, boosting_smote, "-"),
                            "Random Forest" = c(rf_not_balanced, rf_smote, "-"),
                            "XGBoost" = c(xgboost_not_balanced, xgboost_smote, xgboost_down))
final_results
```


<br>


### 7.5 Classification COPD types







